A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-20 tidyindex.qmd
various small changes

---
title: "A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data"
format:
  tandf-pdf:
    keep-tex: true  
pdf-engine: pdflatex
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: 
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-3813-7155
  - name: Ursula Laa
    affiliation: 
      - name: University of Natural Resources and Life Sciences
        department: Institute of Statistics
        address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
        city: Vienna
        country: Austria
    orcid: 0000-0002-0249-6439
  - name: Nicolas Langrené
    affiliation: 
      - name: BNU-HKBU United International College
        department: Department of Mathematical Sciences
        address: 2000 Jintong Road, Tangjiawan, Zhuhai
        city: Zhuhai, Guangdong
        country: China
    orcid: 0000-0001-7601-4618
  - name: Patricia Menéndez
    affiliation:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0003-0701-6315
abstract: |
  Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making. While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their value and variation in values. Here we discuss a modular data pipeline recommendation to assemble indexes. It is universally applicable to index computation and allows investigation of index behavior as part of the development procedure. One can compute with different the index with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess an index's robustness. The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on a country's ranking; and how to calculate bootstrap confidence intervals for XXX index. The methods are supported by a new R package, called `tidyindex`.
keywords:
  - indexes
  - data pipeline
  - software design
  - uncertainty
  - decision-making
bibliography: bibliography.bib  
editor_options: 
  chunk_output_type: console
---

@@

Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making. They serve as critical tools across the natural and social sciences. Examples include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, and the Human Development Index. In environmental science climate indexes are produced by major monitoring centers, like the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early detection of natural disasters.  In economics, indexes provide insight into market trends through combining prices of a basket of goods and services. In social sciences, indexes are used to monitor human development, gender equity, or university quality. The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared.

@@

There is also some attention being given to the diagnosis of indexes, and incorporation of uncertainty. @jones_vulnerability_2007 investigates the methodological choices made in the development of indexes for assessing vulnerable neighborhoods. @uncertainty describes incorporating uncertainty estimates and conducting sensitivity analysis on composite indexes. @tate_social_2012, similarly, makes a comparative assessment of social vulnerability indexes based on uncertainty estimation and sensitivity analysis. (XXX Something about Ursula's colleagues paper here too.)

@@

From reviewing this literature, and in the process of developing methods for making it easier to work with multivariate spatiotemporal data, it seems possible to think about indexes in a more organised, cohesive and standard manner. Actually, it seems that the area could benefit from a *tidy* approach.

@@

```{r fig-scale-var-trans-compare}
#| fig-cap: "Comparison of the module scaling (green) and variable transformation (orange). While both modules change the variable range, scaling maintains the same distributional shape, which is not the case with variable transformation. "
#| fig-height: 8
#| fig-width: 12
set.seed(123)
x <- rgamma(n = 1000, shape = 1, scale = 2)

trans_df <- tibble(origin = x, 
       `z-score` = (x - mean(x)/ sd(x)), 
       qudratic = x^2,
       log = log(x),
       `square root` = sqrt(x),
       `cubic root` = x^(1/3),
       minmax = (x - min(x))/ (max(x) - min(x)),
       boxcox = (x^0.5 - 1)/0.5,
       centering = x - mean(x)
       ) %>% 
  mutate(id = row_number(),
         ) %>% 
  pivot_longer(-id, names_to = "var", values_to = "value") %>% 
  mutate(category = ifelse(
    var %in% c("origin", "z-score", "minmax", "centering"),
    "Scaling", "Var. Trans.")
    )

latex_df <- tibble(
  var = c("origin", "z-score", "qudratic", "log", "square root", "cubic root",
          "minmax", "boxcox", "centering"),
  latex = c(
           r"($x$)",
           r"($\frac{x - \bar{x}}{sd(x)}$)",
           r"($x^2$)",
           r"($log(x)$)",
           r"($\sqrt{x}$)",
           r"($x^{1/3}$)",
           r"($\frac{x - x_{min}}{x_{max} - x_{min}}$)",
           r"($\frac{x^{0.5} - 1}{0.5}$)",
           r"($x - \bar{x}$)"
         )
) %>% 
  left_join(trans_df %>% group_by(var) %>% summarise(max = max(value)))

trans_df %>% 
  ggplot() +
  geom_density(aes(x = value, color = category, y = after_stat(scaled))) + 
  geom_label(data = latex_df, 
             aes(x = 0.8 * max, y = 0.8, label = TeX(latex, output = "character")), 
             parse = TRUE) + 
  facet_wrap(vars(var), scales = "free") + 
  scale_color_brewer(palette = "Dark2", name = "module") + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

@@

```{r fig-compute-spatial}
#| fig-cap: "Spatial distribution of Standardized Precipitation Index (SPI-12) in Queensland, Australia during two major flood and drought events: 2010/11 and 2019/20. The map shows a continuous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 January, likely due to the increased rainfall in February from the meteorological record."
#| fig-width: 6
queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
  rmapshaper::ms_simplify(keep = 0.02)
queensland_map %>%
  ggplot() +
  geom_sf(fill = "transparent", color = "grey50", linewidth = 0.4) +
  geom_point(
    data = idx %>%
      filter(.idx == "spi", .scale == 12) %>%
      filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
               ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )),
    aes(x = long, y = lat, color = .index)) +
  colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
  theme_void() +
  facet_wrap(vars(ym), ncol = 6)
```

@@

render(
  dt[,5:8], planned2_tour(b), dev = "png",
  display_idx(half_range = 2,cex = 4, label_cex = 3.5, col = "black",
              panel_height_ratio = c(4,1),
              axis_bar_lwd = c(rep(3, 3), 7),
              axis_bar_col = c(rep("#000000", 3), "red"),
              axis_bar_label_cex = c(rep(2.5, 3), 3),
              axis_bar_label_col = c(rep("#000000", 3), "red"), 
              axis_label_cex = 3, frame_x_pos = 0.16, 
              axis_var_cex = 2.5,
              label = dt$country, abb_vars = FALSE),
  width = 900, height = 1400, apf = 1/20, frames = 120,
  here::here("figures/idx-tour/idx-tour-%03d.png")
)

@@

TODO: increase font for variable names

## Confidence interval [TO BE CHANGED]

@@

In SPI, the distribution fit step fits the gamma distribution to the aggregated precipitation series separately for each month. This results in  31 or 32 points, from 1990 Jan to 2022 Apr, for estimating each set of distribution parameters. To account for this sampling uncertainty with these samples, bootstrapping is used to generate replicates of the aggregated series. In the `tidyindex` package, this bootstrap sampling is activated when the argument `.n_boot` is set to a value other than the default of 1. In the following code, the Standardized Precipitation Index (SPI) is calculated using a time scale of 24. The bootstrap procedure samples the aggregated precipitation (`.agg`) for 100 iterations ( `.n_boot = 100`) and then fits the gamma distribution. The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `agument()`.

@@

The confidence interval can then be calculated using the quantile method from the bootstrap samples. @fig-conf-interval presents the 80% and 95% confidence interval for the Texas post office station, in Queensland, Australia. From the start of 2019 to 2020,  the majority of the confidence intervals lie below the extreme drought line (SPI = -2), suggesting a high level of certainty that the Texas post office is suffering from a drastic drought. The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantifying the drought/ flood severity in extreme events.

```{r fig-conf-interval}
#| fig-width: 8
#| fig-height: 2
#| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI-24) for the Texas post office station, in Queensland, Australia. A bootstrap sample of 100 is taken from the aggregated pricipitation series to estimate gamma parameters and to calculate the index. The dashed line at SPI = -2 represents an extreme drought as defined by the SPI. Most parts of the confidence intervals from 2019 to 2020 sit below the extreme drought line (-2) and are relatively wide compared to other time periods. This suggests that while it is certain that the Texas post office is suffering from a drastic drought, there is considerable uncertainty in quantifying its severity, given the extremity of the event."
res <- queensland %>% 
  filter(name == "TEXAS POST OFFICE") %>% 
  init(id = id, time = ym) %>% 
  aggregate(.var = prcp, .scale = 24) %>% 
  dist_fit(.dist = gamma(), .var = .agg, .n_boot = 100) %>% 
  augment(.var = .agg)

>>>>

---
title: "A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data"
format:
  tandf-pdf:
    keep-tex: true  
pdf-engine: pdflatex
author:
  - name: H. Sherry Zhang
    affiliations:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-7122-1463
    email: huize.zhang@monash.edu
  - name: Dianne Cook
    affiliation: 
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0002-3813-7155
  - name: Ursula Laa
    affiliation: 
      - name: University of Natural Resources and Life Sciences
        department: Institute of Statistics
        address: Gregor-Mendel-Straße 33, 1180 Wien, Austria
        city: Vienna
        country: Austria
    orcid: 0000-0002-0249-6439
  - name: Nicolas Langrené
    affiliation: 
      - name: BNU-HKBU United International College
        department: Department of Mathematical Sciences
        address: 2000 Jintong Road, Tangjiawan, Zhuhai
        city: Zhuhai, Guangdong
        country: China
    orcid: 0000-0001-7601-4618
  - name: Patricia Menéndez
    affiliation:
      - name: Monash University
        department: Department of Econometrics and Business Statistics
        address: 29 Ancora Imparo Way, Clayton
        city: Melbourne, Victoria
        country: Australia
    orcid: 0000-0003-0701-6315
abstract: |
  Indexes are useful for summarizing multivariate information into single metrics for monitoring, communicating, and decision-making. While most work has focused on defining new indexes for specific purposes, more attention needs to be directed towards making it possible to understand index behavior in different data conditions, and to determine how their structure affects their values and variation in values. Here we discuss a modular data pipeline recommendation to assemble indexes. It is universally applicable to index computation and allows investigation of index behavior as part of the development procedure. One can compute indexes with different parameter choices, adjust steps in the index definition by adding, removing, and swapping them to experiment with various index designs, calculate uncertainty measures, and assess indexes' robustness. The paper presents three examples to illustrate the pipeline framework usage: comparison of two different indexes designed to monitor the spatio-temporal distribution of drought in Queensland, Australia; the effect of dimension reduction choices on the Global Gender Gap Index (GGGI) on countries' ranking; and how to calculate bootstrap confidence intervals for the Standardized Precipitation Index (SPI). The methods are supported by a new R package, called `tidyindex`.
keywords:
  - indexes
  - data pipeline
  - software design
  - uncertainty
  - decision-making
bibliography: bibliography.bib  
editor_options: 
  chunk_output_type: console
---

@@

Indexes are commonly used to combine and summarize different sources of information into a single number for monitoring, communicating, and decision-making. They serve as critical tools across the natural and social sciences. Examples include the Air Quality Index, El Niño-Southern Oscillation Index, Consumer Price Index, QS University Rankings, and the Human Development Index. In environmental science, climate indexes are produced by major monitoring centers, like the United States Drought Monitor and National Oceanic and Atmospheric Administration, to facilitate agricultural planning and early detection of natural disasters.  In economics, indexes provide insight into market trends through combining prices of a basket of goods and services. In social sciences, indexes are used to monitor human development, gender equity, or university quality. The problem is that every index is developed in its own unique way, by different researchers or organizations, and often indexes designed for the same purpose cannot easily be compared.

@@

There is also some attention being given to the diagnosis of indexes, and incorporation of uncertainty. @jones_vulnerability_2007 investigates the methodological choices made in the development of indexes for assessing vulnerable neighborhoods. @uncertainty describes incorporating uncertainty estimates and conducting sensitivity analysis on composite indexes. @tate_social_2012 and @tate_uncertainty_2013, similarly, make a comparative assessment of social vulnerability indexes based on uncertainty estimation and sensitivity analysis. @laimighofer_how_2022 studies five uncertainty sources (record length, observation period, distribution choice, parameter estimation method, and GOF-test) of drought indexes.

@@

From reviewing this literature, and in the process of developing methods for making it easier to work with multivariate spatio- temporal data, it seems possible to think about indexes in a more organised, cohesive and standard manner. Actually, it seems that the area could benefit from a *tidy* approach.

@@

```{r fig-scale-var-trans-compare}
#| fig-cap: "Comparison of the module scaling (green) and variable transformation (orange). While both modules change the variable range, scaling maintains the same distributional shape, which is not the case with variable transformation. "
#| fig-height: 6
#| fig-width: 8
set.seed(123)
x <- rgamma(n = 1000, shape = 1, scale = 2)

trans_df <- tibble(origin = x, 
       `z-score` = (x - mean(x)/ sd(x)), 
       qudratic = x^2,
       log = log(x),
       `square root` = sqrt(x),
       `cubic root` = x^(1/3),
       minmax = (x - min(x))/ (max(x) - min(x)),
       boxcox = (x^0.5 - 1)/0.5,
       centering = x - mean(x)
       ) %>% 
  mutate(id = row_number(),
         ) %>% 
  pivot_longer(-id, names_to = "var", values_to = "value") %>% 
  mutate(category = ifelse(
    var %in% c("origin", "z-score", "minmax", "centering"),
    "Scaling", "Variable Transformation")
    )

latex_df <- tibble(
  var = c("origin", "z-score", "qudratic", "log", "square root",
      "cubic root","minmax", "boxcox", "centering"),
  latex = c(
           r"($x$)",
           r"($\frac{x - \bar{x}}{sd(x)}$)",
           r"($x^2$)",
           r"($log(x)$)",
           r"($\sqrt{x}$)",
           r"($x^{1/3}$)",
           r"($\frac{x - x_{min}}{x_{max} - x_{min}}$)",
           r"($\frac{x^{0.5} - 1}{0.5}$)",
           r"($x - \bar{x}$)"
         )
) %>% 
  left_join(trans_df %>% group_by(var) %>% summarise(max = max(value)))

trans_df %>% 
  ggplot() +
  geom_density(aes(x = value, color = category, y = after_stat(scaled))) + 
  geom_label(data = latex_df, 
             aes(x = 0.8 * max, y = 0.8, label = TeX(latex, output = "character")), 
             parse = TRUE) + 
  facet_wrap(vars(factor(var, levels = c(
      "origin", "centering", "minmax","z-score", "boxcox", "cubic root",
      "log", "qudratic", "square root"))), scales = "free") + 
  scale_color_brewer(palette = "Dark2", name = "module") + 
  theme_bw() + 
  theme(legend.position = "bottom")
```

@@

```{r fig-compute-spatial}
#| fig-cap: "Spatial distribution of Standardized Precipitation Index (SPI-12) in Queensland, Australia during two major flood and drought events: 2010/11 and 2019/20. The map shows a continuous wet period during the 2010/11 flood period and a mitigated drought situation, after its worst in 2019 December and 2020 January, likely due to the increased rainfall in February from the meteorological record."
#| fig-width: 6
#| fig-height: 3
queensland_map <- ozmaps::abs_ste %>% filter(NAME == "Queensland") %>% 
  rmapshaper::ms_simplify(keep = 0.02)
queensland_map %>%
  ggplot() +
  geom_sf(fill = "transparent", color = "grey50", linewidth = 0.4) +
  geom_point(
    data = idx %>%
      filter(.idx == "spi", .scale == 12) %>%
      filter(((year(ym) %in% c(2010, 2019) & month(ym) >= 10) )|
               ((year(ym) %in% c(2011, 2020) & month(ym) <= 3) )),
    aes(x = long, y = lat, color = .index)) +
  colorspace::scale_color_continuous_divergingx(palette = "Geyser", rev = TRUE, name = "Index") + 
  theme_void() +
  facet_wrap(vars(ym), ncol = 6)
```

@@

render(
  dt[,5:8], planned2_tour(b), dev = "png",
  display_idx(half_range = 2,cex = 4, label_cex = 3.5, col = "black", 
              panel_height_ratio = c(4,1),
              axis_bar_lwd = c(rep(3, 3), 7),
              axis_bar_col = c(rep("#000000", 3), "red"),
              axis_bar_label_cex = c(rep(2.5, 3), 3),
              axis_bar_label_col = c(rep("#000000", 3), "red"), 
              axis_label_cex = 3, frame_x_pos = 0.16,  frame_cex = 4,
              axis_var_cex = 3.6, 
              label = dt$country, abb_vars = FALSE),
  width = 900, height = 1400, apf = 1/20, frames = 120,
  here::here("figures/idx-tour/idx-tour-%03d.png")
)

@@

## Decoding uncertainty through the wisdom of the crowd

@@

In SPI, the distribution fit step fits the gamma distribution to the aggregated precipitation series separately for each month. This results in  31 or 32 points, from January 1990 to April 2022, for estimating each set of distribution parameters. To account for this sampling uncertainty with these samples, bootstrapping is used to generate replicates of the aggregated series. In the `tidyindex` package, this bootstrap sampling is activated when the argument `.n_boot` is set to a value other than the default of 1. In the following code, the Standardized Precipitation Index (SPI) is calculated using a time scale of 24. The bootstrap procedure samples the aggregated precipitation (`.agg`) for 100 iterations (`.n_boot = 100`) and then fits the gamma distribution. The resulting gamma probabilities are then transformed into normal densities in the normalizing step with `agument()`.

@@

The confidence interval can then be calculated using the quantile method from the bootstrap samples. @fig-conf-interval presents the 80% and 95% confidence interval for the Texas post office station, in Queensland, Australia. From the start of 2019 to 2020,  the majority of the confidence intervals lie below the extreme drought line (SPI = -2), suggesting a high level of certainty that the Texas post office is suffering from a drastic drought. The relatively wide confidence interval, as well as during the excessive precipitation events in 1996-1998 and 1999-2000, suggests a high variation of the gamma parameters estimated from the bootstrap samples and its difficulty to accurately quantifying the drought and flood severity in extreme events.

```{r fig-conf-interval}
#| fig-width: 8
#| fig-height: 2
#| fig-cap: "80% and 95% confidence interval of the Standardized Precipitation Index (SPI-24) for the Texas post office station, in Queensland, Australia. A bootstrap sample of 100 is taken from the aggregated pricipitation series to estimate gamma parameters and to calculate the index. The dashed line at SPI = -2 represents an extreme drought as defined by the SPI. Most parts of the confidence intervals from 2019 to 2020 sit below the extreme drought line and are relatively wide compared to other time periods. This suggests that while it is certain that the Texas post office is suffering from a drastic drought, there is considerable uncertainty in quantifying its severity, given the extremity of the event."
res <- queensland %>% 
  filter(name == "TEXAS POST OFFICE") %>% 
  init(id = id, time = ym) %>% 
  aggregate(.var = prcp, .scale = 24) %>% 
  dist_fit(.dist = gamma(), .var = .agg, .n_boot = 100) %>% 
  augment(.var = .agg)

### 2023-08-20 tidyindex.tex
various small changes

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their value and variation in values. Here we discuss a
modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
with different the index with different parameter choices, adjust steps
in the index definition by adding, removing, and swapping them to
experiment with various index designs, calculate uncertainty measures,
and assess an index's robustness. The paper presents three examples to
illustrate the pipeline framework usage: comparison of two different
indexes designed to monitor the spatio-temporal distribution of drought
in Queensland, Australia; the effect of dimension reduction choices on
the Global Gender Gap Index (GGGI) on a country's ranking; and how to
calculate bootstrap confidence intervals for XXX index. The methods are
supported by a new R package, called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, interior hidden, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

There is also some attention being given to the diagnosis of indexes,
and incorporation of uncertainty. Jones and Andrey (2007) investigates
the methodological choices made in the development of indexes for
assessing vulnerable neighborhoods. Saisana, Saltelli, and Tarantola
(2005) describes incorporating uncertainty estimates and conducting
sensitivity analysis on composite indexes. Tate (2012), similarly, makes
a comparative assessment of social vulnerability indexes based on
uncertainty estimation and sensitivity analysis. (XXX Something about
Ursula's colleagues paper here too.)

@@

From reviewing this literature, and in the process of developing methods
for making it easier to work with multivariate spatiotemporal data, it
seems possible to think about indexes in a more organised, cohesive and
standard manner. Actually, it seems that the area could benefit from a
\emph{tidy} approach.

@@

TODO: increase font for variable names

\hypertarget{confidence-interval-to-be-changed}{%
\subsection{Confidence interval {[}TO BE
CHANGED{]}}\label{confidence-interval-to-be-changed}}

@@

In SPI, the distribution fit step fits the gamma distribution to the
aggregated precipitation series separately for each month. This results
in 31 or 32 points, from 1990 Jan to 2022 Apr, for estimating each set
of distribution parameters. To account for this sampling uncertainty
with these samples, bootstrapping is used to generate replicates of the
aggregated series. In the \texttt{tidyindex} package, this bootstrap
sampling is activated when the argument \texttt{.n\_boot} is set to a
value other than the default of 1. In the following code, the
Standardized Precipitation Index (SPI) is calculated using a time scale
of 24. The bootstrap procedure samples the aggregated precipitation
(\texttt{.agg}) for 100 iterations ( \texttt{.n\_boot\ =\ 100}) and then
fits the gamma distribution. The resulting gamma probabilities are then
transformed into normal densities in the normalizing step with
\texttt{agument()}.

@@

The confidence interval can then be calculated using the quantile method
from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
80\% and 95\% confidence interval for the Texas post office station, in
Queensland, Australia. From the start of 2019 to 2020, the majority of
the confidence intervals lie below the extreme drought line (SPI = -2),
suggesting a high level of certainty that the Texas post office is
suffering from a drastic drought. The relatively wide confidence
interval, as well as during the excessive precipitation events in
1996-1998 and 1999-2000, suggests a high variation of the gamma
parameters estimated from the bootstrap samples and its difficulty to
accurately quantifying the drought/ flood severity in extreme events.

@@

\caption{\label{fig-conf-interval}80\% and 95\% confidence interval of
the Standardized Precipitation Index (SPI-24) for the Texas post office
station, in Queensland, Australia. A bootstrap sample of 100 is taken
from the aggregated pricipitation series to estimate gamma parameters
and to calculate the index. The dashed line at SPI = -2 represents an
extreme drought as defined by the SPI. Most parts of the confidence
intervals from 2019 to 2020 sit below the extreme drought line (-2) and
are relatively wide compared to other time periods. This suggests that
while it is certain that the Texas post office is suffering from a
drastic drought, there is considerable uncertainty in quantifying its
severity, given the extremity of the event.}

@@

@@

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, breakable, interior hidden, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

There is also some attention being given to the diagnosis of indexes,
and incorporation of uncertainty. Jones and Andrey (2007) investigates
the methodological choices made in the development of indexes for
assessing vulnerable neighborhoods. Saisana, Saltelli, and Tarantola
(2005) describes incorporating uncertainty estimates and conducting
sensitivity analysis on composite indexes. Tate (2012) and Tate (2013),
similarly, make a comparative assessment of social vulnerability indexes
based on uncertainty estimation and sensitivity analysis. Laimighofer
and Laaha (2022) studies five uncertainty sources (record length,
observation period, distribution choice, parameter estimation method,
and GOF-test) of drought indexes.

@@

From reviewing this literature, and in the process of developing methods
for making it easier to work with multivariate spatio- temporal data, it
seems possible to think about indexes in a more organised, cohesive and
standard manner. Actually, it seems that the area could benefit from a
\emph{tidy} approach.

@@

\hypertarget{decoding-uncertainty-through-the-wisdom-of-the-crowd}{%
\subsection{Decoding uncertainty through the wisdom of the
crowd}\label{decoding-uncertainty-through-the-wisdom-of-the-crowd}}

@@

In SPI, the distribution fit step fits the gamma distribution to the
aggregated precipitation series separately for each month. This results
in 31 or 32 points, from January 1990 to April 2022, for estimating each
set of distribution parameters. To account for this sampling uncertainty
with these samples, bootstrapping is used to generate replicates of the
aggregated series. In the \texttt{tidyindex} package, this bootstrap
sampling is activated when the argument \texttt{.n\_boot} is set to a
value other than the default of 1. In the following code, the
Standardized Precipitation Index (SPI) is calculated using a time scale
of 24. The bootstrap procedure samples the aggregated precipitation
(\texttt{.agg}) for 100 iterations (\texttt{.n\_boot\ =\ 100}) and then
fits the gamma distribution. The resulting gamma probabilities are then
transformed into normal densities in the normalizing step with
\texttt{agument()}.

@@

The confidence interval can then be calculated using the quantile method
from the bootstrap samples. Figure~\ref{fig-conf-interval} presents the
80\% and 95\% confidence interval for the Texas post office station, in
Queensland, Australia. From the start of 2019 to 2020, the majority of
the confidence intervals lie below the extreme drought line (SPI = -2),
suggesting a high level of certainty that the Texas post office is
suffering from a drastic drought. The relatively wide confidence
interval, as well as during the excessive precipitation events in
1996-1998 and 1999-2000, suggests a high variation of the gamma
parameters estimated from the bootstrap samples and its difficulty to
accurately quantifying the drought and flood severity in extreme events.

@@

\caption{\label{fig-conf-interval}80\% and 95\% confidence interval of
the Standardized Precipitation Index (SPI-24) for the Texas post office
station, in Queensland, Australia. A bootstrap sample of 100 is taken
from the aggregated pricipitation series to estimate gamma parameters
and to calculate the index. The dashed line at SPI = -2 represents an
extreme drought as defined by the SPI. Most parts of the confidence
intervals from 2019 to 2020 sit below the extreme drought line and are
relatively wide compared to other time periods. This suggests that while
it is certain that the Texas post office is suffering from a drastic
drought, there is considerable uncertainty in quantifying its severity,
given the extremity of the event.}

@@

\leavevmode\vadjust pre{\hypertarget{ref-laimighofer_how_2022}{}}%
Laimighofer, Johannes, and Gregor Laaha. 2022. {``How Standard Are
Standardized Drought Indices? {Uncertainty} Components for the {SPI} \&
{SPEI} Case.''} \emph{Journal of Hydrology} 613 (October): 128385.
\url{https://doi.org/10.1016/j.jhydrol.2022.128385}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-tate_uncertainty_2013}{}}%
---------. 2013. {``Uncertainty {Analysis} for a {Social}
{Vulnerability} {Index}.''} \emph{Annals of the Association of American
Geographers} 103 (3): 526--43.
\url{https://doi.org/10.1080/00045608.2012.700616}.
